The Unicode Cookbook \linebreak for Linguists

### 2018-03-06 book/reviews/responses_reviewer1.txt

# Responses to reviewer 1

*** Our responses prefixed with "***" (which is sometimes also used to delimit the comments from the reviewer for easier answering).

*** We kindly thank the reviewer -- we know reviewing (especially a book) is a time consuming endeavor and we feel that their comments have helped us substantially improve the cookbook. (Note that chapter numbers mentioned by the reviewer may have changed, e..g we split chapter 1 into chapters 1 and 2, as suggested.)

===========
As stated in the preface, the book not only targets linguists, but also
programmers. This review is written by a programmer from the perspective
of a programmer.

I want to start out saying that this is a much needed book. I personally 
often felt a need to educate linguists about unicode and the representation 
of text in computers and was hoping for a reference with enough detail *and* 
the proper language fit for linguists. The "Unicode Cookbook for Linguists"
feels like it is this reference, so I highly recommend publication of it by
Language Science Press.

I also welcome and second the call to linguists to take IPA and Unicode for
what they are: Bona fide, best effort standards with well-defined use cases.
The invitation to treat these standards as useful tools in a pragmatic way
rather than complaining about their perceived shortcomings is repeated several
times throughout the book - but bears repeating, because reluctance to
accept and adhere to common standards in language documentation, not a rare 
trait among linguists, has prevented the field from more meaningfully employ
quantitative methods of language comparison.

The first chapter of the "Unicode Cookbook for Linguists" provides a good
introduction to the problem space of written language and how to compare it.
Thus, it describes the motivation and problems for standardization efforts.
The last section of chapter 1, "1.4 The Unicode approach" may fit better as
first section in the second chapter, at least in my mental model of the 
outline of the book as

1. The problem space
2. Proposed solution from an IT perspective
3. Proposed solution from a linguistic perspective
4. Identification and description of the "impedance mismatch" between the two
5. Recommendation how to bridge the gap between the two standards

*** We have separated out the Unicode Approach subsection of chapter 1 into its own chapter (2) with these subsections that appear in the TOC:

- Background
- The Unicode Standard
- Character encoding system
- Grapheme clusters

Anyway, chapters 2 and 3 introduce the two most important attempts at
standardizing written language for linguists working on computers: The
Unicode standard and the International Phonetic Alphabet. These two standards
are described with enough history and detail to appreciate the results, but
also highlighting - with the benefit of hindsight - the shortcomings.

*** Thanks!

Chapter 2, "Unicode Pitfalls", is arranged as a list of common problems when
working with Unicode. Even just looking at the some section names, 
"Characters are not glyphs", "Characters are not graphs", "Blocks", "Names",
hints at the main problem of using Unicode being one of terminology.

In particular for programmers or researchers using a computing environment like
python or R, this is certainly the case, because many of these environments -
being mainly used and developed in the English-speaking world - get away with
not distinguishing two things, namely "sequences of code points" and "text
encoded as bit patterns". Now looking up "Character encoding" on Wikipedia [1]
we learn that the term "encoding" is used for all levels - code points or
bit patterns. But in programming, the standard terminology seems to have
settled on reserving "encoding" for the bit-pattern level, while using terms
like "string" or "unicode text" for sequences of Unicode code points [2].

So while it may be technically correct to say

    The Unicode Standard is a CHARACTER ENCODING SYSTEM (page 13)

this sounds a bit odd to the Unicode-aware programmer, who would rather say
"UTF-8 is a Unicode Encoding system". So the solution of the Unicode Standard
for the problem of many mutually conflicting encoding systems was to introduce
a new layer of indirection between character and bit pattern - the code point.
Unfortunately, the Unicode Consortium itself seems to shy away from clearer
*new* terminology, when it compares Unicode to encoding systems (not saying
Unicode is one, though), but does not coin a new term for what Unicode actually
is, but rather defining it purely functional as "Unicode provides a unique number 
for every character" [3].
It's hard to tell whether attempting to coin such a term could have succeeded
with this book, but it is a bit sad, that the blurry terminology used everywhere
is perpetuated here. Still, the authors should try not to distract or confuse
the reader as with footnote 10 on page 13, where the short excursion into
numeral systems seems to suggest, that Unicode code point U+0070 is always
"encoded" as bit pattern 1110 0000.

*** We appreciate this keen comment from the programmer's perspective. In attempt to not confuse the targeted audiences (ordinary working linguists, potential and amateur programmers), we've added a footnote to this sentence "The Unicode Standard is a CHARACTER ENCODING SYSTEM (page 13)", which we think clarifies the terminological issue for advanced readers:

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems. The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e. the code point, which can be encoded differently by different encoding systems}

We have also removed the footnote: "Hexadecimal (base-16) 0070 is equivalent to decimal
(base-10) 112..."

Furthermore, we thought it pertinent to add an additional Pitfall: Encodings, which we think clarifies the reviewer's points about the terminological mismatch of "encoding". We put this section before the file formats pitfall.

***

Section "Pitfall: File formats" may profit most from a more transparent
terminology, distinguishing levels of encoding, rather than talking about how
texts "appear inside some kind of computer file". This section should also
mention that line breaks are actually part of the text, and as such also
covered by the Unicode standard. Again regarding line breaks, the 
recommendation "that everybody use this encoding [only LF] whenever possible"
(page 31) seems less well motivated (and specific) than most other 
recommendations in the book. Some common formats, e.g. CSV (as specified by 
RFC 4180 [4]), specify "CRLF" as line break. Thus, the perceived "strong 
tendency" towards "LF" may be just that.

*** See above.

One Unicode pitfall which may be worth adding is the problem with incomplete
implementations of the Unicode Standard. This is a problem most big-enough
standards suffer from, e.g. SQL, XML, XSLT. So although the Unicode Standard
mandates that comparison of Unicode text should always be done using normalized
text, this is not how the equality operator "==" is implemented for Unicode
text in the Python programming language.

*** We added a very brief Pitfall that addresses this point (and shamelessly took the examples from the reviewer). Thanks!

Pitfall: Incomplete implementations

Another pitfall that we encounter when using the Unicode Standard is its incomplete implementation in different standards and programming languages, e.g.\ SQL, XML, XLST, Python. For example, although the Unicode Standard mandates that the comparison of Unicode text be done using normalized text, this is not the csae with the equality operator ``=='' in Python. Furthermore, it is not always transparent what the operating system or specific software applications do when text is being copied and pasted. For example, copy and pasting the character sequence \uni{0061} \textsc{latin small letter a} <a> and \uni{0301} \textsc{combining acute accent} <\dia{0301}>, visually <á>, into the text editor TextWrangler leaves the sequence decomposed as two characters. But when pasting the decomposed sequence into RStudio, and other software programs, the sequence becomes precomposed as \uni{00E1} \textsc{latin small letter a with acute}, i.e.\ <á>.

***

Chapter 3 introduces "The International Phonetic Alphabet". As with Unicode,
the authors stress the characterization of IPA as a standard, which makes some
premises and is guided by some principles, but overall tries to solve an
interoperability problem in a pragmatic way. I.e. when you want to use text
on a computer, you should use Unicode; if you want to compare written language
across languages, you should use IPA.

It may be worth highlighting the parallel between "IPA Numbers" and "Unicode 
Code Points", i.e. the fact, that IPA and Unicode came up with the solution of
an additional layer of indirection between symbols/characters and encoding
(on bit-pattern level); or the parallel between SAMPA/X-SAMPA and the multitude
of pre-Unicode encodings for text. The fact that the authors added enough
historical background about both IPA and Unicode makes it possible to relate
these parallels in time, and prevent dismissal of attempts like SAMPA as
misguided - because now we have Unicode.

*** We added a note about this to the last paragraph in Chapter 3:

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and the Unicode Consortium came up with the solution of an additional layer of indirection between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful as an input method for IPA in ASCII and required by some software.

***

Again, the authors should take care to keep terminology as consistent as 
possible, e.g. in a passage like

    […] the computer-coding convention for the IPA should be independent of 
    computer environments and formats (e.g. ASCII), i.e. the IPA Number was
    not meant to be implemented directly in a computer encoding (page 41)

"computer-coding convention", "format" and "computer encoding" seem to refer
to the same thing, i.e. encoding on bit-pattern level. In particular listing
ASCII as an example of a format makes a programmer wonder whether this new
term is significant, because up to this point, ASCII was referred to as a
character encoding.

*** We've updated this passage:

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats, 
i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.

Chapter 4, "IPA meets Unicode", describes what it means for IPA that Unicode
has become *the* standard for handling text in a computing environment. We
have learned so far that 
- we have to use Unicode to use text with a computer, and
- we have to use IPA to compare written language.
Thus, chapter 4 gives recommendations how to compare written language with a 
computer. As with chapter 2, this chapter is organized as list of pitfalls,
augmented with a section on recommendations. While it may seem overly verbose
or boring that most of the pitfalls contain somewhat extensive lists of
symbols/characters/Unicode code blocks, I think this is overall the right
approach: As the autors show, making IPA work with Unicode is largely a
matter of spelling out special cases, and how to deal with them. It seems
important, though, to have the tables starting at page 70 available in a
computer readable format, and have the book point to the location of these
files.

*** Nice suggestion! We have put these tables online here in machine-readable format: https://github.com/unicode-cookbook/cookbook/tree/master/book/tables

Chapter 5 introduces "Orthography profiles" as a tool to specify writing
systems in such a way that transliteration (e.g. to a suitable subset of IPA)
becomes simple and transparent; thus, a tool to get from the Unicode domain
of handling text with a computer to the IPA domain of comparing written
language with a computer. Having used orthography profiles myself in projects,
I can already testify to their usefulness.

Like a "real" programmer would do, whenever a chapter contains a section called
"Formal specification", I'll skip right ahead to it.

File Format:

A1: "ideally using […]" is not exactly language you would like to see in a
specification; instead the specification should follow RFC 2119 [5], 
"Key words for use in RFCs to Indicate Requirement Levels". So in this case,
I guess an orthography profile must be normalized following NFC (or NFD, if 
specified in the metadata). Specifying a convention for line endings and BOM
seems overly strict, since most computing environments will transparently
handle both alternatives: E.g. using the Python programming language, decoding
a file using the encoding "utf-8-sig" will strip away a BOM, if present, and
reading a file in text mode will interpret both "LF" and "CRLF" as line
endings.

*** Thank you for the link to RFC 2219 -- we have changed our wording in accordance with their keywords. For example, we've updated this paragraph to:

A1. A profile is a unicode UTF-8 encoded text file normalized following NFC (or NFD if speci ed in the metadata) that includes the information pertinent to the orthography.^2

[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of our experience in working with linguistic data in this format. A keen reviewer notes that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives. For example, using Python a  file can be decoded using the encoding “utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.

A2: "tabs and newline […] will lead to problems", thus they should be
explicitely disallowed!

*** Agreed! We have updated this to:

A2. A profile is a CSV file with an obligatory header line. A minimal pro le must have a single column with the header “Grapheme”. For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.  e actual ordering of the columns is unimportant. Cells should not con- tain tabs or newlines.  e header list must be delimited in the same way as the rest of the  le’s contents. Each record must be kept on a separate line. Separate lines with comments are not allowed. Comments that belong to specific lines must be put in a separate column of the CSV file, e.g.~add a column called \textsc{comments}.

Note we also merged in A4 into A2, previously:

A4. Separate lines with comments are not allowed. Comments that be- long to speci c lines will have to be put in a separate column of the CSV  le, e.g. add a column called comments.

***

A3: The metadata file seems to be highly under-specified. At the very least,
there should be a list of recognized tags providing semantics for the data.
But since machine-readability is a requirement for the file anyway, a format
like JSON-LD [6], or even more appropriate, the JSON-LD dialect specified in
"Metadata Vocabulary for Tabular Data" [7]  may be a better option, because it
allows easy inclusion of Dublin Core [8] metadata - as well as a specification
of the exact CSV dialect used for the actual orthography profile.
It should be noted that JSON-LD metadata is the choice for datasets conforming
to the Cross-Linguistic Data Formats [9] standard.

*** Agreed. In practice, we haven't actually used these metadata files. We've updated this paragraph to "SHOULD" via RFC 2119 and we RECOMMEND they follow the practices suggested by the reviewer and implemented in CLDF:

A3. Metadata should be added in a separate UTF-8 text file using the JSON-DL dialect speci ed in Metadata Vocabulary for Tabular Data.^ 3  This
metadata format allows easy inclusion of Dublin Core metadata,^4 which should be used to specify information about the orthographic description in the orthography pro le.  e orthography pro le metadata should min- imally include: (i) author, (ii) date, (iii) title of the pro le, (iv) a stable lan- guage identi er encoded in BCP 47/ISO 639-3 of the target language of the pro le, and (v) bibliographic data for resource(s) that illustrate the or- thography described in the pro le. Further, (vi) the tokenization method and (vii) the Unicode normalization form used should be documented here (see below). In accordance with the Metadata Vocabulary for Tabular Data, the metadata’s  lename must consist of the orthography pro le’s  lename appended with “-metadata.json”.^5

[3] https://www.w3.org/TR/tabular-metadata/
[4] http://dublincore.org/
[5] JSON-DL metadata is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, see: http://cldf.clld.org/.

A7: "If regular expressions are used, then […]": The specification does not
state how a processing application can determine if that is the case - other
than through additional user input. So specifiying whether regular expressions
are used should be required in the metadata file. In general, allowing regular
expressions as input is somewhat problematic, because regular expression
syntax as well as functionality is somewhat under-specified across regex
implementations in different computing environments. The closest to a standard
in this area might be "POSIX Basic Regular Expressions" [10], but these do not
include lookahead and lookbehind functionality as requred by B3. So allowing
regular expressions as part of orthography profiles may lead to different
results of applying the same profile to the same data on different platforms -
which clearly goes against the idea of this specification.

*** TODO

Overall, it seems somewhat strange to sacrifice
CSV standards conformance (no quoting of cell values, "LF" line endings, 
tab as separator) in order to achieve better read-/writeability by humans. This
puts a bigger burden on the authors in terms of specification (they can no 
longer just point to FC 4180) and it makes it harder to use tools like 
LibreOffice Calc or MS Excel to edit Orthography profiles.

*** Agreed (SM). We have made the formal specification lacker (which of course put the onus the software developers):

A1. A profile is a unicode UTF-8 encoded text file normalized following NFC (or NFD if speci ed in the metadata) that includes the information pertinent to the orthography.2

A2. A profile is a CSV file with an obligatory header line. A minimal pro le must have a single column with the header “Grapheme”. For any ad- ditional columns, the name in the header of the CSV  le must be speci ed.  e actual ordering of the columns is unimportant. Cells should not con- tain tabs or newlines.  e header list must be delimited in the same way as the rest of the  le’s contents. Each record must be kept on a separate line.

[2] See Section 3.12 in which we suggest to use NFC, no-BOM and LF line breaks because of the pitfalls they avoid. A keen reviewer notes, however, that specifying a convention for line endings and BOM is overly strict because most computing environments (now) transparently handle both alternatives. For example, using Python a  le can be decoded using the encoding 	“utf-8-sig”, which strips away the BOM (if present) and reads an input full in text mode, so that both line feed variants “LF” and “CRLF” will be stripped.

Implementation:

B1: As noted for A7, to make the basic handling possible in an automated way,
there must be a way to figure out that the profile is "basic".

*** TODO

B2: If - as stated - "ordering […] might be crucial", it must be specified, i.e.
members a class specified as context must be tried in order of occurrence in the
profile.

*** TODO

B4: I would rather spin the functionality to suggest an optimal ordering of rules
off to a separate tool, and limit the specification to apply rules in order of
occurrence. This will lead to simpler implementation and more transparent 
relation of input and output.

*** TODO

Language like

    The actual implementation of the profile on some text-string will function 
    as follows: (page 83)

sounds a bit weird in a technical specification. In general, the presentation
may benefit by adopting the conventions as used for describing algorithms, i.e.
specifying
- input
- output
- a series of steps to get from input to output

*** TODO

I did not review the Python and R implementations, and suggest to spin these off
to a more suitable format, e.g. Jupyter Notebooks [11].

[1] https://en.wikipedia.org/wiki/Character_encoding
[2] https://docs.python.org/3/howto/unicode.html#encodings
[3] http://www.unicode.org/standard/WhatIsUnicode.html
[4] https://www.ietf.org/rfc/rfc4180.txt
[5] https://www.ietf.org/rfc/rfc2119.txt
[6] https://json-ld.org/
[7] https://www.w3.org/TR/tabular-metadata/
[8] http://dublincore.org/
[9] http://cldf.clld.org/
[10] http://www.regular-expressions.info/posix.html
[11] http://jupyter.org/

*** In addition to having this material in the PDF version of the cookbook, we have spun off the implementations (and installation instructions) into Jupyter notebooks (https://github.com/unicode-cookbook/cookbook/issues/22):

For installation of the R and Python libraries (in the README):
https://github.com/unicode-cookbook/recipes https://github.com/unicode-cookbook/recipes

R basics:
https://github.com/unicode-cookbook/recipes/blob/master/Basics/R.ipynb https://github.com/unicode-cookbook/recipes/blob/master/Basics/R.ipynb

### 2018-03-06 book/reviews/responses_reviewer1.txt

Python basics:
https://github.com/unicode-cookbook/recipes/blob/master/Basics/python.ipynb https://github.com/unicode-cookbook/recipes/blob/master/Basics/python.ipynb

The four other recipes are also now in Jupyter notebooks.

### 2018-03-06 book/chapters/implementation.tex
fixes the tabs in the code should have the same length, so that the headers and the columns align
these changes almost complete the remaining reviews

To illustrate the practical application of orthography profiles as defined in
the previous Chapter~\ref{orthography-profiles}, we have implemented two 
versions of the specifications presented there: one in Python and one
in R.

These two libraries have rather different implementation histories,
and we are not 100\% sure that they will in all situations give the same
results. We do however provide a test suite for each implementation. 
Users should refer to the tests cases and to the documentation in each release 
for specifics about each implementation. Note also that function names differ 
in the two implementations.

Note also that the performance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as proof of concept and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple and sometimes 
somewhat abstract examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles.

This chapter is split into two sections, one for the Python implementation, the 
other for R. Each section has a number of subsections that ...

In addition to the material presented here to get users started, we maintain 
several case studies that use each OP implementation in action. These use cases 
are an the accompanying Github repository available with this book at: 
\url{https://github.com/unicode-cookbook/recipes/}.

% TODO: point to the case studies
% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.

\section{Python}
\label{python-implementations}

The Python package is available both as a command line interface (CLI) 
and as an appplication programming interface (API).

@@

\subsection*{Application programming interface}
The \texttt{segments} API can be accesses by importing 
the package into Python a script. Here is an example of 
how to import the libraries, create a tokenizer object, 
tokenize a string and create a profile.

\begin{lstlisting}[language=python]
>>> from __future__ import unicode_literals, print_function
>>> from segments import Profile, Tokenizer
>>> t = Tokenizer()
>>> t('abcd')
'a b c d'
>>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
{'Grapheme': 'cd', 'mapping': 'y'})
>>> print(prf)
mapping Grapheme
ab  x
cd  y
>>> t = Tokenizer(profile=prf)
>>> t('abcd')
'ab cd'
>>> t('abcd', column='mapping')
'x y'
\end{lstlisting}

\section{R}
\label{r-implementation}

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Section~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

There are various additional options for rule ordering implemented. Please check
the help description in R \texttt{help(tokenize)} for more details on the
possible rule ordering specificiations. In summary, there are four different 
ordering options, that can also be combined:

@@

It is important to realize that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied becuase a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specifed order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules' is rather easy to predict.

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard-coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
For example, we are adding an extra line to the profile from above, resulting in
the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
graphemes are based on regular expressions so you can also use regular
expressions in the description of the context. For such contextually specified
graphemes to be included in the graphemic parsing we have to specify the option
\texttt{regex = TRUE}. This contextually specified grapheme should actually be
handled first, so we could try \texttt{ordering = NULL}. However, we can also
explicitly specifiy that rules with contextual information should be applied
first by using \texttt{ordering = "context"}. That gives the right results for
this toy example.

@@

Normally, the separator specified should not occur in the data. If it does,
unexpected things might happen, so consider removing the chosen seperator from
your strings first. However, there is also an option \texttt{sep.replace} to
replace the separator with something else. When \code{sep.replace} is specified,
this mark is inserted in the string at those places where the separator occurs.
Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
spaces between graphemes and replacing spaces in the input string by hashes in
the output string.

@@

>>>>

\section{Overview}
To illustrate the practical applications of orthography profiles, as defined in Chapter~\ref{orthography-profiles}, we have implemented two versions of the specifications presented there: one in Python\footnote{\url{https://pypi.python.org/pypi/segments}} and one in R.\footnote{\url{https://github.com/cysouw/qlcData}} This chapter introduces these two implementations, and gives practical step-by-step guidelines for installing and using them. Various simple and sometimes somewhat abstract examples will be discussed to show the different options available, and to illustrate the intended usage of orthography profiles.

Note that our two software libraries have rather different implementation histories, thus they may not give the same results in all situations. We do, however, provide a test suite for each implementation aimed at making sure that the results covered by those test cases are identical. Users should refer to the tests and to the documentation in each release for specifics regarding each implementation. Due to the different naming conventions in Python and R, function names differ across the two implementations. The performance with larger datasets also differs per implementation. In sum, the two software libraries should be considered as proof of concept and not as the final word on the practical application of the specifications above. In our own experience, the current versions are sufficiently fast and stable to be useful for academic practice (e.g.\ checking data consistency, or analyzing and transliterating small to medium sized data sets), but they should probably not be used for full-scale industry applications without some adaptations.

In Section \ref{installing-python-and-r}, we first explain how to install Python\footnote{\url{https://www.python.org/}} and R.\footnote{\url{https://www.r-project.org/}} In Sections \ref{python-implementations} and \ref{r-implementation}, we discuss our Python and R software packages, respectively. In addition to the material presented here to get users started, we maintain several case studies that use orthography profiles in action. For convenience, we make these use cases available as Jupyter Notebooks\footnote{\url{http://jupyter.org/}} in our Github repository of code recipes, which accompanies this book online.\footnote{\url{https://github.com/unicode-cookbook/}} In the final section in this chapter, we briefly describe these additional use-cases and point the user to the accompanying code online.

\section{How to install Python and R}
\label{installing-python-and-r}
When one encounters problems installing software, or bugs in programming code, search engines are your friend! Installation problems and incomprehensible error messages have typically been encountered and solved by other users. Try simply copying and pasting the output of an error message into a search engine; the solution is often already somewhere online. We are fans of Stack Exchange\footnote{\url{https://stackexchange.com/}} -- a network of question-and-answer websites -- which are extremely helpful in solving issues regarding software installation, bugs in code, etc.

Searching the Web for ``install r and python'' returns numerous tutorials on how to set up your machine for scientific data analysis. Note there is no single correct setup for a particular computer or operating system. Both Python and R are available for Windows, Mac, and Unix operating systems from the Python and R project websites. Another option is to use a so-called package manager, i.e.\ a software program that allows the user to manage software packages and their dependencies. On Mac, we use Homebrew,\footnote{\url{https://brew.sh/}} a simple to install (via the Terminal App) free and open source package management system. Follow the instructions on the the Homebrew website and then use Homebrew to install R and Python (as well as other software packages such as Git and Jupyter Notebooks).

Alternatively for R, RStudio\footnote{\url{https://www.rstudio.com/}} provides a free and open source integrated development environment (IDE). This application can be downloaded and installed (for Mac, Windows and Unix) and it includes its own R installation and R libraries package manager. For developing in Python, we recommend the free community version of PyCharm,\footnote{\url{https://www.jetbrains.com/pycharm/}} an IDE which is available for Mac, Windows, and Unix.

Once you have R or Python (or both) installed on your computer, you are ready to use the orthography profiles software libraries presented in the next two sections. As noted above, we make this material available online on Github,\footnote{\url{https://github.com/}} a web-based version control system for source code management. Github repositories can be cloned or downloaded,\footnote{\url{https://help.github.com/articles/cloning-a-repository/}} so that you can work through the examples on your local machine. Use your favorite search engine to figure out how to install Git on your computer and learn more about using Git.\footnote{\url{https://git-scm.com/}} In our Github repository, we make the material presented below (and more use cases described briefly in Section \ref{use-cases}) available as Jupyter Notebooks. Jupyter Notebooks provide an interface where you can run and develop source code using the browser as an interface. These notebooks are easily viewed in our Github repository of use cases.\footnote{\url{https://github.com/unicode-cookbook/recipes}}

\section{Python package: segments}
\label{python-implementations}

The Python package \texttt{segments} is available both as a command line interface (CLI) and as an application programming interface (API).

@@

\section{R library: qlcData}
\label{r-implementation}

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Chapter~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

To leave out the Unicode information in
the profile skeleton, use the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

There are various additional options for rule ordering implemented. Please check
the help description in R \texttt{help(tokenize)} for more details on the
possible rule ordering specifications. In summary, there are four different 
ordering options, that can also be combined:

@@

It is important to realize that different ordering of the rules does not have
so-called feeding and bleeding effects as known from finite-state rewrite
rules.\footnote{Bleeding is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. Feeding is the opposite: a
specific rule will only be applied because a previous rule changed the string
already. The interaction of rules with such feeding and bleeding effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be captured in the 
string. All lines in the profile are processed in a specified order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as captured, but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the rules' is rather easy to predict.

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard-coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
For example, we are adding an extra line to the profile from above, resulting in
the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
graphemes are based on regular expressions so you can also use regular
expressions in the description of the context. For such contextually specified
graphemes to be included in the graphemic parsing we have to specify the option
\texttt{regex = TRUE}. This contextually specified grapheme should actually be
handled first, so we could try \texttt{ordering = NULL}. However, we can also
explicitly specify that rules with contextual information should be applied
first by using \texttt{ordering = "context"}. That gives the right results for
this toy example.

@@

Normally, the separator specified should not occur in the data. If it does,
unexpected things might happen, so consider removing the chosen separator from
your strings first. However, there is also an option \texttt{sep.replace} to
replace the separator with something else. When \code{sep.replace} is specified,
this mark is inserted in the string at those places where the separator occurs.
Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
spaces between graphemes and replacing spaces in the input string by hashes in
the output string.

@@

\section{More recipes}
\label{use-cases}

Online we provide several recipes, or use cases, that illustrate the applications of orthography profiles using our implementations in Python and R. \footnote{\url{https://github.com/unicode-cookbook/recipes}} Here we briefly describe these use cases and we encourage users to try them out using Git and Jupyter Notebooks.

First, as we discussed above, we provide a basic tutorial on how to use the Python \texttt{segments}\footnote{\url{https://pypi.python.org/pypi/segments}} and R \texttt{qlcData}\footnote{\url{https://github.com/cysouw/qlcData}} libraries.\footnote{\url{https://github.com/unicode-cookbook/recipes/tree/master/Basics}} This use case simply shows the basic functions of each library.

The two use cases using the Python \textt{segments} package include a tutorial on how to segment graphemes in IPA text:

\begin{itemize}
	\item https://github.com/unicode-cookbook/recipes/tree/master/JIPA
\end{itemize}

\noindent and an example of how to create an orthography profile to tokenize fieldwork data from a large comparative wordlist.

\begin{itemize}
	\item https://github.com/unicode-cookbook/recipes/tree/master/Dogon
\end{itemize}

\noindent The JIPA use case uses excerpts from the Northwind and the Sun passages from the Illustrations of the IPA published in the Journal of the International Phonetic Alphabet. Thus the use case shows how a user might tokenize IPA proper. The Dogon use case uses fieldwork data from the Dogon languages of Mali language documentation project.\footnote{\url{http://dogonlanguages.org/}} This use case illustrates how a user might tokenize fieldwork data from numerous linguists using different transcription practices by defining these practices with an orthography profile to make the output unified and comparable.

The two use cases using the R \texttt{qlcData} library include a use case for tokenizing wordlist data from the Automated Similarity Judgment Program (ASJP):\footnote{\url{http://asjp.clld.org/}}

\begin{itemize}
	\item https://github.com/unicode-cookbook/recipes/tree/master/ASJP
\end{itemize}

\noindent and for tokenizing a corpus of text in Dutch orthography:

\begin{itemize}
	\item https://github.com/unicode-cookbook/recipes/tree/master/Dutch
\end{itemize}

\noindent The ASJP use case shows how to download the full set of ASJP wordlists, to combine them into a single large CSV file, and to tokenize the ASJP orthography. The Dutch use case uses the 10K corpus for Dutch (``nld'') from the Leipzig Corpora Collection,\footnote{\url{http://corpora2.informatik.uni-leipzig.de/download.html}} which is then cleaned and tokenized with an orthography profile that captures the intricacies of Dutch orthography.
